Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-11 2024pipit/Main.tex
ok
time-triggered example
tr
extraction

\ifx\authoranonymous\relax
\newcommand{\GITHUBURL}{(anonymised)}
\else
\newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
\fi

@@

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate imperative code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  This translation to imperative code preserves types by construction; the proof that the imperative code preserves semantics is ongoing.
\end{abstract}

@@

Big TODOs:
\begin{itemize}
  \item example: trigger-fetch from ttcan;
  \item transition systems and translation;
\end{itemize}

\input{section/01-Introduction.tex}

\input{section/20-Motivation.tex}
\input{section/30-Core.tex}
\input{section/40-Transition.tex}
\input{section/50-Extraction.tex}
\input{section/60-Evaluation.tex}
\input{section/70-Related.tex}
\input{section/80-Conclusion.tex}

>>>>

\ifx\authoranonymous\relax
\newcommand{\GITHUBURL}{(anonymised)}
\newcommand{\COFFEEURL}{(anonymised)}
\else
\newcommand{\GITHUBURL}{\url{https://github.com/songlarknet/pipit}}
\newcommand{\COFFEEURL}{\url{https://youtu.be/6IybbQFPOl8}}
\fi

@@

\begin{abstract}
  Reactive languages such as Lustre and Scade are used to implement safety-critical control systems; proving such programs correct and having the proved properties apply to the compiled code is therefore equally critical.
  We introduce Pipit, a small reactive language embedded in \fstar{}, designed for verifying control systems and executing them in real-time.
  Pipit includes a verified translation to transition systems; by reusing \fstar{}'s existing proof automation, certain safety properties can be automatically proved by k-induction on the transition system.
  Pipit can also generate executable code in a subset of \fstar{} which is suitable for compilation and real-time execution on embedded devices.
  The executable code is deterministic and total and preserves the semantics of the original program.
\end{abstract}

@@

\input{section/01-Introduction.tex}
\input{section/20-Motivation.tex}
\input{section/30-Core.tex}
\input{section/35-Primitive-Tables.tex}
\input{section/40-Transition.tex}
\input{section/50-Extraction.tex}
\input{section/60-Evaluation.tex}
\input{section/70-Related.tex}
\input{section/80-Conclusion.tex}

### 2024-01-11 2024pipit/section/01-Introduction.tex
extraction

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics; we also prove that the imperative code preserves the semantics.

@@

\begin{itemize}
  \item we motivate our work, which allows us to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that gives the advantages of using a step-function such as quantifier-free combination, with the ability to encode relations such as underspecified rely-guarantee contracts (\autoref{s:transition});
  \item we identify the invariants and lemmas required to prove preservation of the dynamic semantics to an abstract transition system suitable for verifying properties (\autoref{s:transition:proof});
  \item similarly, we identify the invariants and lemmas required to prove that the executable transition system preserves the original semantics (\autoref{s:extraction:proof});
  \item finally, we evaluate Pipit by implementing and partially-verifying a time-triggered CAN bus driver (\autoref{s:evaluation}).
\end{itemize}

>>>>

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this paper we introduce Pipit\footnote{Implementation available at \GITHUBURL}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
To verify programs, Pipit translates its expression language to a transition system for k-inductive proofs, which is verified to be an abstraction of the original semantics.
To execute programs, Pipit can generate executable code, which is total and semantics-preserving.

@@

\begin{itemize}
  \item we motivate our work, which allows us to combine manual and automated proofs of reactive systems with a strong specification language (\autoref{s:motivation});
  \item we introduce Pipit, a minimal reactive language that supports rely-guarantee contracts and properties; crucially, proof obligations are annotated with a status --- \emph{valid} or \emph{deferred} --- allowing proofs to be delayed until more is known of the program context (\autoref{s:core});
  \item we describe a \emph{checked semantics} for Pipit, which is parameterised by the property status; after checking deferred properties, programs can be \emph{blessed}, and their properties lifted to valid status (\autoref{s:core:checked});
  \item we describe an encoding of transition systems that can express under-specified rely-guarantee contracts as functions, rather than relations, which results in simpler transition systems (\autoref{s:transition});
  \item we identify and mechanise the invariants and lemmas required to prove that the abstract transition system is an abstraction of the original semantics (\autoref{s:transition:proof});
  \item similarly, we verify that the executable transition system preserves the original semantics (\autoref{s:extraction});
  \item finally, we evaluate Pipit by implementing a time-triggered CAN bus driver, which we have partially verified (\autoref{s:evaluation}).
\end{itemize}

### 2024-01-11 2024pipit/section/20-Motivation.tex
causality etc
extraction

Pipit is an embedded domain-specific language.
The program above is really syntactic sugar for an \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable.
We will discuss the details of the core language in \autoref{s:core}, but for now we will show the source program with some minor embedding details omitted.

@@

The system has strict timing requirements, which restrict how triggers can be defined.
In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
Additionally, the schedule must allow sufficient time for the driver to skip over any disabled triggers.
Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2 instead.
However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per time-mark, and it takes two steps to reach trigger 2 from the start of the array.

@@

\subsection{Instantiating lemmas and defining contracts}

@@

We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 refers to the lemma taking one streaming argument (the index).
% The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
% The $\textit{pose}$ function is a helper function implemented using the @check@ and @bless@ functions from earlier.
% The property we state here is that if the trigger at the current index is active, then there is a sufficient time gap to be able to reach the next active trigger after the current index.
We then state an invariant as a deferred property.
Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if the environment always ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.

@@

% \subsection{Program logics and dependent types}
% \TODO{compare to refinement types, verification conditions, etc here, or wait to \autoref{s:related-work}?}

% Although Pipit is embedded in \fstar{}, which itself supports full dependent types and refinement types, we have opted to for a comparatively weakly-typed language.
% Our stream type \emph{stream $\tau$} denotes a (finite prefix of a) stream of elements of simple type $\tau$, but says nothing about how the stream evolves over time.

% We could consider an alternative approach that was closer to an imperative program logic with preconditions and postconditions.
% For example, we could consider a richer type $\stream \tau~ \{ \rawbind{x}{\phi} \}$ which then the result, of type $\tau$, always satisfies stream postcondition $\phi$.
% We could imagine specifying count_when with such a type:

% \begin{tabbing}
%   @val@ count_when ($\textit{max}$: $\NN$) ($\textit{inc}$: stream $\BB$): stream $\{ \top \} ~\NN~ \{ \rawbind{c}{0 \le c \le max} \}$\\
% \end{tabbing}

% An alternative would be to use refinement types \cite{chen2022synchronous} or some kind of program logic with pre- and post-conditions.

% We believe that such type-based approaches are useful for composing programs together, but 
%  the nested-recursive nature of Lustre programs makes such approaches cumbersome.
% In our system, the recursive combinator has type $@rec@: (\stream \alpha \to \stream \alpha) \to \stream \alpha$.
% Consider instead, if we had a refined stream $\stream \alpha \{ \psi \}$

>>>>

Pipit is an embedded domain-specific language.
The program above is really syntactic sugar for an \fstar{} program that takes a natural number and constructs a Pipit core expression with a free boolean variable.
We will discuss the details of the core language in \autoref{s:core}, but for now we focus on the source program with some minor embedding details omitted.

@@

The system has strict timing requirements which restrict how triggers can be defined.
In this example, each trigger has a unique time; in general, trigger times can overlap, but they need to be enabled on distinct cycles.
Additionally, the schedule must allow sufficient time for the driver to skip over the disabled triggers.
Concretely, we could postpone trigger 1 to send message @B@ at time-mark 2, as triggers 1 and 2 have distinct cycles.
However, we could not bring forward trigger 2 to send message @C@ at time-mark 1: the driver can only process one trigger per tick, and it takes two steps to reach trigger 2 from the start of the array.

@@

\subsection{Instantiating lemmas and defining contracts}
\label{s:motivation:contract}

@@

We use the $\textit{pose}_1$ helper function to lift the \emph{lemma_can_reach_next} lemma to a streaming context and instantiate it; the subscript 1 indicates that the lemma is being applied to one streaming argument (the index).
% The lemma requires a non-streaming cycle and a non-streaming index; we apply it to the static cycle and the combinator lifts the application to the streaming index.
% The $\textit{pose}$ function is a helper function implemented using the @check@ and @bless@ functions from earlier.
% The property we state here is that if the trigger at the current index is active, then there is a sufficient time gap to be able to reach the next active trigger after the current index.
We then state an invariant as a deferred property.
Informally, the invariant states that, either the current active trigger is not late, or the next active trigger after the current index is in the future and we can reach it in time.

With the explicitly instantiated lemma, we can prove the streaming invariant by straightforward induction on the transition system.
To help make proofs of the rest of the system more compositional, we also abstract over the details of the trigger-fetch mechanism by introducing a rely-guarantee contract for \emph{trigger_fetch}.
The contract we state is that if the environment ensures that the time doesn't skip --- that is, we are called once per microsecond --- then we guarantee that we never encounter a late trigger.

@@

### 2024-01-11 2024pipit/section/50-Extraction.tex
ok
extraction

Pipit can generate executable code which is suitable for real-time execution on embedded devices.
The code extraction is implemented in a currently-unverified transform that takes a deeply embedded representation of a Pipit expression and generates a shallow imperative representation of the program.
\fstar{} can generate C code from a subset of the language called \lowstar{}~\cite{protzenko2017verified}; the result of our translation to imperative code fits in this subset.
During code extraction, we use \fstar{}'s tactic support~\cite{martinez2019meta} to fully normalise the translation to imperative code, conceptually similar to staged compilation.

% For the simple examples in \autoref{ss:tut:bibo}, we probably could have proved the properties directly on the generated code.
% In the future, however, we intend to support the abstraction of streams through \emph{contracts}, which require a relational semantics rather than a functional one.

% noextract
% let expr = Sugar.run2 Pump.controller'

% noextract
% let system: Pipit.Exec.Exp.xexec expr =
%   XX.exec_of_exp expr

% [@@(Tac.postprocess_with XL.tac_extract)]
% let reset = XL.mk_reset system

% [@@(Tac.postprocess_with XL.tac_extract)]
% let step (inp: input) = XL.mk_step system (inp.estop, (inp.level_low, ()))
% This allows us to generate C code which executes in bounded space on embedded devices.

>>>>

Pipit can generate executable code which is suitable for real-time execution on embedded devices.
The code extraction uses a variation of the abstract transition system described in \autoref{s:transition}, with two main differences to ensure that the result is executable without relying on the environment to provide values for the free context.
Contracts are straightforward to execute by using the body of the contract rather than abstracting over the implementation.

To execute recursive expressions $\xrec{x}{e} : \tau$, we require an arbitrary value of type $\tau$ to seed the fixpoint, as described in \autoref{s:core:causality}.
We first call the step function to evaluate $e$ with $x$ bound to $\bot_\tau$.
This step call returns the correct value, but the updated state is invalid, as it may refer to the bottom value.
To get the correct state, we call the step function again, this time with $e$ bound to $v$.

This translation to transition systems is verified to preserve the original semantics; however, it can duplicate work by requiring two calls to the step function.
To extract the above transition systems, we use \fstar{}'s normalisation-by-evaluation to generate step functions that update their mutable state in-place and which \lowstar{} can extract to C.
This normalisation strategy inlines the two step functions and is often able to remove the duplicate work.

Unfortunately, our current approach is also unsuitable for generating imperative array code, as our shallowly-embedded pure transition system requires pure arrays.
In the future, we intend to address array computations and the above work duplication by generating an explicit deeply-embedded imperative program.
We currently generate a monolithic step function for each Pipit program; an embedded foreign-function interface similar to that used by Accelerate~\cite{clifton2014embedding} may also enable separate-compilation of programs.

% \item staging: similarity to \cite{gallego2021w};
